---
title: "Clustering iris data"
output: 
  html_document:
    toc: true
    number_sections: true
    keep_rmd: true
---


```{r setup, include=FALSE} 
is_sol = FALSE
knitr::opts_chunk$set(message=FALSE)
```


# Iris  data

```{r}
head(iris)
```
The `iris` data is a classical dataset that comes with R. Read the documentation of the dataset. The dataset contains features of 50 sample of each of three varieties of the iris flower. 

In this case we know the variety (that is the cluster), and the task is to investigate how well clustering performs in identifying these varieties.

Make pairwise plots of all variables. Make sure that the varieties are shown in different colours. Based on such plots, how well do you anticipate the clustering algorithm to work?

```{r}
library(GGally)
library(ggplot2)
## your code

# Sample data (iris dataset)
df <- iris  # Replace with your own dataframe

# Pairwise plot of all variables with different colors for the variety
ggpairs(df, aes(colour = Species))
```


Define dataset with the numerical values only:
```{r}
iris_numeric <- iris[,1:4]
head(iris_numeric)
```

Run a clustering algorithm on this dataset. Should data be scaled before doing the clustering?

```{r, eval=F}
# without scaling
hc <- hclust(dist(iris_numeric))

print(head(iris_numeric))


df_scaled <- scale(iris_numeric)

print(head(df_scaled))

hc <- hclust(dist(df_scaled))

print(hc)

# answer yes it should be scaled before otherwised Big numeric numbers could screwer the culster. But it is not allways the case

# scaling makes the number between -2 to 2
```


Extract 10 clusters and make a cross tabulation of the clusters and the species in the `iris` dataset. Here, `table()` is a good friend.

```{r, eval=F}

cluster <- cutree(hc, k = 10)

table(cluster, iris$Species)

# Plot the dendrogram
plot(hc, labels = iris$Species, main = "Dendrogram of Iris Clustering", xlab = "", sub = "")

```



Experiment with extracting 2, 3 and 4 clusters from data and cross tabulate these the species. Based on this, how well do will you say that clustering works for these data? Are you surprised?


Create scatterplots where you label points according to  the case `k=3`:

```{r}
kVar = 3

cluster <- cutree(hc, k = kVar)

table(cluster, iris$Species)

# Plot the dendrogram
plot(hc, labels = iris$Species, main = "Dendrogram of Iris Clustering", xlab = "", sub = "")


# Create boxes around the 5 clusters
rect.hclust(hc, k = kVar, border = 2:6)
```




<!-- This file has been generated automatically from clustering_itvest_24_ex.rmd -->

<!-- time: 2024-09-20 07:07:22.852225 do not edit manually -->

