---
title: "Relational data and database exercises"
format: html
---

## Load packages and data

```{r tidy-setup, message = FALSE}
library(tidyverse)
library(nycflights13)
```

## Exercise 1: Average arrival delay for each carrier

What is the average arrival delay on flights for each carrier?

Show the results together with both carrier code and carrier name

(*Hint:* Use an inner join of `flights` and `airlines` and appropriate `group_by()` and `summarise()`.)

```{r}
average_delay_by_carrier <- flights |> 
  inner_join(airlines, by = c("carrier" = "carrier")) |>          # Inner join Carrier from table flights and airlines
  group_by(carrier, name) |>                                      # Group by both carrier code and carrier name
  summarise(average_arr_delay = mean(arr_delay, na.rm = TRUE)) |> # Calculate average arrival delay, ignoring NAs
  arrange(desc(average_arr_delay))                                # Arrange by descending order of delay

# View the result
average_delay_by_carrier

```


## Exercise 2: Flights without records in `planes`

For each carrier, what's the number of flights that don't have matching records in `planes`?

Show results together with carrier code and name as before

*Hint:* We need three different tables and two different joins (have a look at `anti_join()` for the first part).

```{r}
?flights

# Find flights without matching records in the planes table
Flights_not_in_planes <- flights |>
  anti_join(planes, by = "tailnum")

# Join with airlines to get carrier names and count the unmatched flights per carrier
Flights_without_records <- Flights_not_in_planes |> 
  inner_join(airlines, by = "carrier") |> 
  group_by(carrier, name) |> 
  summarise(n_flights = n())

# View the result
head(Flights_without_records)
```

## Exercise 3: Number of flights by carrier and destination

What is the number of flights for each carrier and destination?

Show the name of both the carrier and airport, not the id codes, and sorted by airline and airport for clarity.

*Hint:* We need to join three different tables and notice that two tables have a variable `name` with different meanings so they will appear with two different suffixes (you can rename them if you like).

```{r}
# 
number_of_fligths <- flights |> 
  inner_join(airlines, by = "carrier") |> 
  inner_join(airports, by = c("dest" = "faa"))
  
```

