---
title: "Relational data and database exercises"
format: html
---

## Load packages and data

```{r tidy-setup, message = FALSE}
library(tidyverse)
library(nycflights13)
```

## Exercise 1: Average arrival delay for each carrier

What is the average arrival delay on flights for each carrier?

Show the results together with both carrier code and carrier name

(*Hint:* Use an inner join of `flights` and `airlines` and appropriate `group_by()` and `summarise()`.)

```{r}

```


## Exercise 2: Flights without records in `planes`

For each carrier, what's the number of flights that don't have matching records in `planes`?

Show results together with carrier code and name as before

*Hint:* We need three different tables and two different joins (have a look at `anti_join()` for the first part).

```{r}

```

## Exercise 3: Number of flights by carrier and destination

What is the number of flights for each carrier and destination?

Show the name of both the carrier and airport, not the id codes, and sorted by airline and airport for clarity.

*Hint:* We need to join three different tables and notice that two tables have a variable `name` with different meanings so they will appear with two different suffixes (you can rename them if you like).

```{r}

```

