---
title: "Logistic regression for budworms"
format: 
  html:
    toc: true
    number_sections: true
  pdf:
    toc: true
    number_sections: true
---

# Budworm data

```{r}
library(tidyverse)
```

The following vectors contain the information about the outcome of 12 different
experimental settings (log-dose of insecticide, sex of budworm) for 20 budworms 
in each setting (240 budworms in total):

```{r}
ldose <- rep(0:5, 2)
dead <- c(1, 4, 9, 13, 18, 20, 0, 2, 6, 10, 12, 16)
sex <- factor(rep(c("M", "F"), c(6, 6)))
```

## Exercise 1.1:

Make a `tibble` called `budworms` containing the data (possibly adding the
number of alive budworms if needed) which can be used to fit a logistic regression:

```{r}
budworms <- tibble()
## Insert your code
```


## Exercise 1.2:

Make dataset called `budworms2` where the male and female budworms are aggregated for each dose:

```{r}
budworms2 <- budworms |> group_by() |> summarise()
## Insert your code
```


## Exercise 1.3:

Make a scatterplot for the empirical probability of death depending on `ldose` (hint: you need to calculate the probability before plotting):
```{r, eval=FALSE}
budworms2 |> ## possibly add a mutate here
  ggplot(aes()) +
  geom_point()
```


# Simple logistic regression

## Exercise 2.1:

Fit a simple logistic regression with `ldose` as explanatory variable ignoring 
`sex` and summarise the resulting model with `broom::tidy()`. 

* Is the effect of `ldose` statistically significant?

* What is the interpretation of estimated coefficient for `ldose`?


```{r}
# Model fit and summary code here
library(broom)
```


## Exercise 2.2

Find an estimate of LD50 in this model and give an approximate 95% confidence interval for LD50.
```{r, eval=FALSE}
car::deltaMethod(model, "YOUR_CODE")
```




## Exercise 2.3

Add predicted probabilities of death to the data and make a plot showing both 
the predicted and empirical probabilities.

```{r}
library(modelr)
## Your code
```



# Multiple logistic regression

## Exercise 3.1

Fit a multiple logistic regression with additive effects of `ldose` and `sex` on 
the (logit) probability of death.

```{r, eval=FALSE}
model_add <- glm("YOUR_CODE")
tidy(model_add)
```


## Exercise 3.2

Fit a multiple logistic regression with interaction between effects of `ldose` 
and `sex` on the (logit) probability of death. Is the interaction effect statistically significant?

```{r, eval=FALSE}
model_int <- glm("YOUR_CODE")
tidy(model_int)
```





## Exercise 3.3

Choose one of the models from 3.1 and 3.2 and find an estimate of LD50 in this 
model and give an approximate 95% confidence interval for LD50.

```{r, eval=FALSE}
## *For male budworms in the interaction model:*
car::deltaMethod(model_int, "YOUR_CODE")
```



## Exercise 3.4

Use `pivot_longer()` to make a tidy dataset with 4 columns: `sex`, `ldose`, `status` (dead/alive), `n` (number of budworms).
```{r}
tidy_budworms <- budworms # More commands here
```


Use `uncount()` to expand the dataset to represent the 240 individual budworms 
and then add a predicted status to each budworm according to the model using the 
natural threshold of 50%.
```{r}
indiv_budworms <- tidy_budworms # More commands here
```


Summarise the (mis)classification.


## Exercise 3.5 (optional)

Is "expanding" / "unfolding" / "uncounting" the data as in the previous exercise 
necessary to find the total (mis)classification? 

How could you find the (mis)classification summary directly with the aggregated 
data?

<!-- This file has been generated automatically from logistic_budworm.qmd -->

<!-- time: 2024-10-03 22:27:49.003551 do not edit manually -->

