---
title: "Data Science Exam: Part 2"
author: "Flemming Christensen"
date: "2024-10-6"
format:
  html:
    embed-resources: true
---

# Libraries
```{r}
library(doBy)
library(ggplot2)
library(reshape2)
library(rpart)
library(randomForest)
```


# Part 1: Regression

# Data
This part uses the `personality` dataset from the R package `doBy` with recordings of 32 variables describing personality characteristics for 240 people. (see help file for `?personality`)

We focus on the variable `easygon` as the response variable. We split the data in training and test data as follows:

```{r}
dat <- doBy::personality
set.seed(101) # for reproducibility
i_train <- sample(nrow(dat), .5*nrow(dat))
train <- dat[i_train,]
test <- dat[-i_train,]
```


# Exercise 1
Use `response_plot()` from the **development version of** `doBy` to visualize the relation between the response and the other variables in the data.

```{r, fig.width=10, fig.height=20}
response_plot(dat, easygon ~ ., geoms = c(geom_jitter(alpha = .6, width = .4, height = .4)), ncol = 5)
```


# Exercise 2
Specify a number of prediction models – at least one of each of the following:

1. linear with stepwise selection (`lm`/ `step`)
2. regression tree (`rpart`)
3. random forest (`randomForest`)

## General observations
predictors that seems to be significant predictors from the response plot are following.

Very strong: relaxed - laidbck.
strong: coopera - approvn.

Lets see if that holds up later on.

## 1. linear with stepwise selection (`lm`/ `step`)

```{r, results="hide"}
# Fit a linear model
model <- lm(easygon ~ ., data = train)

# Perform stepwise selection based on AIC
stepwise_model <- step(model, direction = "both")

summary(stepwise_model)
```

## 2. regression tree (`rpart`)

```{r, fig.width=7, fig.height=7}
# Fit a regression tree model
tree_model <- rpart(easygon ~ ., data = train, method = "anova")

# Visualize the tree
plot(tree_model)
text(tree_model)
```

## 3. random forest (`randomForest`)

```{r, fig.width=7, fig.height=7}
# Fit a random forest model
rf_model <- randomForest(easygon ~ ., data = train, ntree = 500)

# Print model details
print(rf_model)

# Visualizes how the model's error rate changes with the number of trees.
plot(rf_model, main = "Random Forest Error Rates")

# Plots the importance of each variable, helping identify which predictors are most influential. "Only for random forrest"
varImpPlot(rf_model, main = "Variable Importance in Random Forest")
```

# Exercise 3
For **one of** the linear models show the estimated coefficients and comment **briefly** on the output (for example, comment on the sign of the estimates).

```{r}
summary(stepwise_model)
```


# Exercise 4
Compare the performance of the fitted models by predicting the values of `easygon` for the test data and show a plot comparing the predictions to the observations from the test data. Also make a table of the RMSE for each model.

```{r}
# Predict on test data
pred_step <- predict(model, newdata = test)
pred_rpart <- predict(tree_model, newdata = test)
pred_rf <- predict(rf_model, newdata = test)

# Extract actual values from the test set
actual <- test$easygon

# Function to calculate RMSE
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

# Calculate RMSE for each model
rmse_step <- rmse(actual, pred_step)
rmse_rpart <- rmse(actual, pred_rpart)
rmse_rf <- rmse(actual, pred_rf)

# Create a table to compare RMSE
rmse_table <- data.frame(
  Model = c("Stepwise Linear", "Regression Tree", "Random Forest"),
  RMSE = c(rmse_step, rmse_rpart, rmse_rf)
)

# Print RMSE table
print(rmse_table)
```

```{r}
# Combine the actual and predicted values into a single data frame for plotting
plot_data <- data.frame(
  Observed = actual,
  Stepwise = pred_step,
  RegressionTree = pred_rpart,
  RandomForest = pred_rf
)

# Melt the data frame to long format for ggplot
plot_data_melt <- melt(plot_data, id.vars = "Observed", 
                       variable.name = "Model", value.name = "Predicted")

# Create the plot
ggplot(plot_data_melt, aes(x = Observed, y = Predicted, color = Model)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  labs(title = "Observed vs Predicted Values of easygon",
       x = "Observed easygon", y = "Predicted easygon") +
  facet_wrap(~Model)
```


# Part 2: Classification
The dataset `diabetes.csv` is about risk factors for diabetes and contains 768 observations of 9 variables. The response variable is `Outcome` which is binary. The goal is to predict `Outcome` using the other variables. A description of the dataset can be found at https://www.kaggle.com/datasets/mathchi/diabetes-data-set/data

# Exercise 1
Read in and clean the data by detecting unexpected values of some variables, replace the values with `NA` and finally make a data set with only the complete cases.

# Exercise 2
Plot the response against each of the explanatory variables/features and add a smooth line as a very rough indication of how the probability of diabetes depends on the individual feature (note the response has to be numeric for the smooth line to work).

# Exercise 3
Ensure Outcome is a factor and split the data in a training and test set as follows:

```{r}
set.seed(202) # for reproducibility
i_train <- sample(nrow(dat), .5*nrow(dat))
train <- dat[i_train,]
test <- dat[-i_train,]
```

Specify a number of classification models and report the corresponding confusion matrix for the test data. You must include at least one of each of the following:

1. logistic regression with stepwise selection (`glm` / `step`)
2. regression tree (`rpart`)
3. random forest (`randomForest`)
4. linear/quadratic discriminant analysis

## 1. logistic regression with stepwise selection (`glm` / `step`)

## 2. regression tree (`rpart`)

## 3. random forest (`randomForest`)

## 4. linear/quadratic discriminant analysis

## Project dependencies

```{r}
sessionInfo()
```