---
title: "Data Science Exam: Part 2"
author: "Flemming Christensen"
date: "2024-10-6"
format:
  html:
    embed-resources: true
---

# Libraries
```{r}
library(doBy)
library(ggplot2)
```


# Part 1: Regression

# Data
This part uses the `personality` dataset from the R package `doBy` with recordings of 32 variables describing personality characteristics for 240 people. (see help file for `?personality`)

We focus on the variable `easygon` as the response variable. We split the data in training and test data as follows:

```{r}
dat <- doBy::personality
set.seed(101) # for reproducibility
i_train <- sample(nrow(dat), .5*nrow(dat))
train <- dat[i_train,]
test <- dat[-i_train,]
```


# Exercise 1
Use `response_plot()` from the **development version of** `doBy` to visualize the relation between the response and the other variables in the data.

```{r, fig.width=10, fig.height=20}
response_plot(dat, easygon ~ ., geoms = c(geom_jitter(alpha = .6, width = .4, height = .4)), ncol = 5)
```


# Exercise 2
Specify a number of prediction models – at least one of each of the following:

1. linear with stepwise selection (`lm`/ `step`)
2. regression tree (`rpart`)
3. random forest (`randomForest`)

## 1. linear with stepwise selection (`lm`/ `step`)

## 2. regression tree (`rpart`)

## 3. random forest (`randomForest`)

# Exercise 3
For **one of** the linear models show the estimated coefficients and comment **briefly** on the output (for example, comment on the sign of the estimates).


# Exercise 4
Compare the performance of the fitted models by predicting the values of `easygon` for the test data and show a plot comparing the predictions to the observations from the test data. Also make a table of the RMSE for each model.

# Part 2: Classification
The dataset `diabetes.csv` is about risk factors for diabetes and contains 768 observations of 9 variables. The response variable is `Outcome` which is binary. The goal is to predict `Outcome` using the other variables. A description of the dataset can be found at https://www.kaggle.com/datasets/mathchi/diabetes-data-set/data

# Exercise 1
Read in and clean the data by detecting unexpected values of some variables, replace the values with `NA` and finally make a data set with only the complete cases.

# Exercise 2
Plot the response against each of the explanatory variables/features and add a smooth line as a very rough indication of how the probability of diabetes depends on the individual feature (note the response has to be numeric for the smooth line to work).

# Exercise 3
Ensure Outcome is a factor and split the data in a training and test set as follows:

```{r}
set.seed(202) # for reproducibility
i_train <- sample(nrow(dat), .5*nrow(dat))
train <- dat[i_train,]
test <- dat[-i_train,]
```

Specify a number of classification models and report the corresponding confusion matrix for the test data. You must include at least one of each of the following:

1. logistic regression with stepwise selection (`glm` / `step`)
2. regression tree (`rpart`)
3. random forest (`randomForest`)
4. linear/quadratic discriminant analysis

## 1. logistic regression with stepwise selection (`glm` / `step`)

## 2. regression tree (`rpart`)

## 3. random forest (`randomForest`)

## 4. linear/quadratic discriminant analysis

## Project dependencies

```{r}
sessionInfo()
```